\ttitle

### 2015-07-26 Chapters/Introduction.tex
Introduction - Multiple edits

Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection. Research is ongoing to assess the role of neoadjuvant therapy and newer treatment modalities such as electroporation in these patients to improve resectability.

\subsubsection{Patient factors}
\subsection{Operative technique}
Pancreaticoduodenectomy is considered one of the most technically challenging operations on the gastrointestinal tract. While the procedure is carried out in a broadly similar fashion in all major centres, there remain some variations in perioperative care as well as some operative steps. The following is a description of the procedure as performed at the West of Scotland Pancreatic Unit.

After a comprehensive preoperative work-up including both assessments of the tumour as well as patient fitness, informed consent was obtained. Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. General anaesthesia with complete muscle relaxation was used in all patients. Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. Antibiotic prophylaxis is administered at induction. While the use of Octreotide, a somatostatin analogue, to reduce the risk of postoperative pancreatic fistula formation is still debated, it was routinely used in all patients at this centre. Octreotide was administered intra-operatively (200 mcg s.c.) and was continued for 5 days postoperatively (200 mcg s.c., t.d.s.).

A roof-top incision was used for access. After assessing the peritoneal cavity for absence of metastatic disease, an early assessment was made for local resectability. This involved complete Kocherisation of the duodenum to assess the retroperitoneum. Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). The rest of the procedure was performed as described extensively elsewhere. The gastrocolic omentum was divided to enter the lesser sac. The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. If less than half the circumference of the SMV or PV was involved, an en-bloc resection was performed with vein repair at the same time. The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. The gastro-duodenal artery was divided. Resection was then completed by dividing the stomach (classical Whipple procedure) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy, PPPD) and transecting the pancreatic neck.

@@

\subsection{Complications}
The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006,emick_hospital_2006,yeo_six_1997, winter_1423_2006,teh_patient_2009,gouma_rates_2000}

\subsubsection{Postoperative pancreatic fistula}
\label{sec:ch_intro_POPF}
Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.\parencite{yeo_six_1997,deoliveira_assessment_2006,bassi_postoperative_2005,winter_biochemical_2007,pratt_risk_2008} 
The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be a result of breakdown or poor healing at the pancreaticojejunostomy/pancreaticogastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis. It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing. Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. Grade B and C fistulae are considered to be clinically significant in that they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage, delayed gastric emptying as well as need for intervention (either radiological or operative) and/or prolonged critical care support.

@@

\subsubsection{Clavien-Dindo classification of complications}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrythmias, pneumonia, pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal sepsis including intra-abdominal sepsis, leakage from the hepaticojejunostomy or the gastrojejunostomy, renal dysfunction, etc. The Clavien-Dindo method grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis.

\input{Tables/c1_clavien_dindo}

@@

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
\subsection{Comorbidity}
Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or chemoradiotherapy.\parencite{sandroussi_sociodemographics_2010}

Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009}

Major pancreatic surgery requires the patient to have adequate physiological reserve to cope with the increased demand during and immediately after surgery. However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery. \parencite{shah_limitations_2012}

\subsection{Risk Stratification}
\todo{Intro - risk stratification}
%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery. Major surgery places a significant physiological stress on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and the postoperative recovery plays a major role in determining short-term outcomes.

Accurate measurement of physiological fitness

\subsection{Static Versus Dynamic Testing}
\todo{Static Versus Dynamic Testing}
Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase. Identification of such high-risk patients allows not only for improved patient selection, but also for risk-stratified, anaesthetic and postoperative critical care. Preoperative risk stratification will also allow for prehabilitation of these patients in an attempt to improve outcomes.

@@

@@

\subsection{Measuring the Anaerobic Threshold}
The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method\parencite{sue_metabolic_1988} or the ventilatory equivalents method.\parencite{beaver_new_1986} Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing.\parencite{society_ats/accp_2003}

@@

\begin{figure}[p]
	\centering
	\includegraphics[width=0.9\textwidth]{Figures/cpet_9panel.pdf}
	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
	\label{fig:cpet_9panel}
\end{figure}

@@

During aerobic exercise, VO2 and VCO2 share a linear relationship as shown in segment A of the graph in figure 1. However, as anaerobic respiration starts to supplement aerobic respiration, VCO2 increases disproportionate to VO2 as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}. This results in a distinct difference in the slope of the initial part of the graph (seg A) and the later part (seg B). The point at which the two slopes intersect is the anaerobic threshold and the V02 at this point in exercise is commonly referred to as the anaerobic threshold, VO2at or simply AT.

\subsubsection{Ventilatory equivalents method}
\todo{Ventilatory equivalents method}

@@

$\dot{V}_{O_2}$ or oxygen uptake is measured breath-by-breath using digital analysis of the inspired and expired gases. This is then averaged, usually over time, to smooth-out any signficant breath-by-breath variation. $\dot{V}_{O_2}$ increases with increasing work load and is influenced by several factors that have a role in the transport and utilisation of oxygen. These may be broadly classified as cardiac, pulmonary, circulatory and tissue factors. Some of the factors are encompassed in the following formula for $\dot{V}_{O_2}$.
\begin{equation} \label{eq:VO2=CaO2xCO}
	\dot{V}_{O_2} = CaO_2 \times Cardiac\ Output
\end{equation}
where $CaO_2$ is $O_2$ content per ml of blood and is defined by,
\begin{equation} \label{eq:CaO2=Hbx1.34xSaO2}
	CaO_2 = Haemoglobin \times 1.34 \times SaO_2
\end{equation}
and
cardiac output, the primary cardiac factor that influences $\dot{V}_{O_2}$, is:
\begin{equation} \label{eq:CO=SVxHR}
	Cardiac\ Output = Stroke\ Volume \times Heart\ Rate
\end{equation}
Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.

@@

\subsection{Oxygen Pulse, $O_2Pulse$}
Oxygen pulse is defined as the oxygen uptake per heart beat.
\begin{equation} \label{eq:O2Pulse=VO2/HR}
	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
\end{equation}
While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. The clinical application of oxygen pulse in surgical patients remains unclear.

@@

\subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
The change in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise. Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. However, as the anaerobic threshold is passed, $\dot{V}_E/\dot{V}_{O_2}$ starts increasing before $\dot{V}_E/\dot{V}_{CO_2}$. This change in direction is yet another method to confirm the anaerobic threshold.  $\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as well as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.

@@

These reports in conjunction with the studies on preoperative inflammation, sepsis, SIRS and CARS emphasise the important role of peri;operative systemic inflammation as a causative factor in postoperative complications and the impact of the 'second-hit' of postoperative complications on long-term survival after curative cancer surgery.

>>>>

Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases. The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced disease and in some patients with metastatic disease. Early reports suggest that some of these patients can undergo potentially curative surgery after such neo-adjuvant treatment.

\subsubsection{Patient factors}

\subsection{Operative technique}
Pancreaticoduodenectomy is considered one of the most technically challenging operations on the gastrointestinal tract. While the procedure is carried out in a broadly similar fashion in all major centres, there remain some variations in perioperative care as well as some operative steps. The following is a description of the procedure as performed at the West of Scotland Pancreatic Unit.

After a comprehensive preoperative work-up including both assessments of the tumour as well as patient fitness, informed consent was obtained. Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. General anaesthesia with complete muscle relaxation was used in all patients. Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. Antibiotic prophylaxis is administered at induction. While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre. Octreotide was administered intra-operatively (200 micrograms subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms subcutaneous injection administered three times a day).

A roof-top incision was made to enter the abdominal cavity. After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability. This involved complete Kocherisation of the duodenum to assess the retroperitoneum. Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). The rest of the procedure was performed as described extensively elsewhere. The gastrocolic omentum was divided to enter the lesser sac. The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time. The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery should the patient require embolisation of this vessel for post-pancreatectomy haemorrhage. The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference. The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. The retroportal lamina was divided preserving the superior mesenteric artery.

@@

\subsection{Complications}
The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000} Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes. The consensus definitions of the International Study Group for Pancreatic Surgery (ISGPS) and the Clavien-Dindo classification systems were used during the course of the studies reported here.

\subsubsection{Postoperative pancreatic fistula}
\label{sec:ch_intro_POPF}

Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.\parencite{yeo_six_1997,deoliveira_assessment_2006,bassi_postoperative_2005,winter_biochemical_2007,pratt_risk_2008} 
The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be a result of breakdown or poor healing at the pancreaticojejunostomy/pancreaticogastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis. It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing. Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying. They may require further intervention (either radiological or operative) and may result in prolonged critical care support.

@@

\subsubsection{Clavien-Dindo classification of complications}
\input{Tables/c1_clavien_dindo}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc. The Clavien-Dindo system grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis. Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.

@@

While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care. The European Study Group for Pancreatic Cancer (ESPAC) have conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect. \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}

The ESPAC-1 trial compared the use of chemotherapy with fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma. The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive demonstrating a clear survival benefit for adjuvant chemotherapy. However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. The ESPAC-3 trial compared the efficacy of 5-FU and folinic acid with gemcitabine and reported that gemcitabine was not superior to 5-FU/folinic acid.

However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma end up either not receiving or not completing adjuvant chemotherapy. \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbdity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
\subsection{Comorbidity}
Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. Major pancreatic surgery requires the patient to have adequate reserve to cope not only with the increased physiological demand during and immediately after surgery but also to cope with any adverse events that may occur.

The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment. \parencite{sandroussi_sociodemographics_2010}

Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery. \parencite{shah_limitations_2012}

\subsection{Risk Stratification}
\todo{Intro - risk stratification}
%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
Physiological fitness or reserve may be defined as the ability of the patient's organ systems to respond appropriately and adequately to the stress of surgery. Major surgery has a significant physiological impact on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes. Therefore, accurate assessment of physiololgical fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions. A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness. Numerous methods including individual tests and composite scores derived from physiological and biochemical variables have been used for risk assessment.

\subsection{Static Versus Dynamic Testing}
\todo{Static Versus Dynamic Testing}

Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase.

@@

\subsection{Origins and physiological basis of cardiopulmonary exercise testing}
The measurement of aerobic fitness has been an important aspect of improving atheletic performance in competitive sports. Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes as well as offer customised training regimens aimed at improving performance at activities that place a signficant stress on the cardiovascular and respiratory systems.

@@

\subsection{Measuring the Anaerobic Threshold}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet_9panel.pdf}
	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
	\label{fig:cpet_9panel}
\end{figure}

The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method\parencite{sue_metabolic_1988} or the ventilatory equivalents method.\parencite{beaver_new_1986} Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing.\parencite{society_ats/accp_2003}

@@

@@

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet_vslope}
	\caption{Determination of $\dot{V}_{O_2}$AT by the V-slope method.}
	\label{fig:cpet_vslope}
\end{figure}

During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}. However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionately to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}. This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B). The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold, $\dot{V}_{O_2}$AT or simply AT.

\subsubsection{Ventilatory equivalents method}
\todo{Ventilatory equivalents method}
The ventilatory equivalents method of determining $\dot{V}_{O_2}$AT involves plotting $\dot{V}_E/\dot{V}_{O_2}$
and $\dot{V}_E/\dot{V}_{CO_2}$ against time, exercise load or $\dot{V}_{O_2}$. The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig. \ref{fig:cpet_9panel}F.

@@

$\dot{V}_{O_2}$ or oxygen uptake is measured breath-by-breath using digital analysis of the inspired and expired gases. This is then averaged, usually over time, to smooth-out any signficant breath-by-breath variation. $\dot{V}_{O_2}$ increases with increasing work load and is influenced by several factors that have a role in the transport and utilisation of oxygen. These may be broadly classified as cardiac, pulmonary, circulatory and tissue factors. Some of the factors are encompassed in the following formula for $\dot{V}_{O_2}$.
\begin{equation} \label{eq:oxygen_consumption}
	\dot{V}_{O_2} = CaO_2 \times Cardiac\ Output
\end{equation}
where $CaO_2$ is $O_2$ content per ml of blood and is defined by,
\begin{equation} \label{eq:CaO2}
	CaO_2 = Haemoglobin \times 1.34 \times SaO_2
\end{equation}
and
cardiac output, the primary cardiac factor that influences $\dot{V}_{O_2}$, is:
\begin{equation} \label{eq:CO=SVxHR}
	Cardiac\ Output = Stroke\ Volume \times Heart\ Rate
\end{equation}
Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.

@@

\subsection{Oxygen Pulse, $O_2Pulse$}
Oxygen pulse is defined as the oxygen uptake per heart beat.
\begin{equation} \label{eq:oxygen_pulse}
	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
\end{equation}
While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. The clinical application of oxygen pulse in surgical patients remains unclear.

@@

\subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
The changes in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise. Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. However, as the anaerobic threshold is passed, $\dot{V}_E/\dot{V}_{O_2}$ starts increasing before $\dot{V}_E/\dot{V}_{CO_2}$. This change in direction is yet another method to confirm the anaerobic threshold.  $\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.

@@

These reports in conjunction with the studies on preoperative inflammation, sepsis, SIRS and CARS emphasise the important role of perioperative systemic inflammation as a causative factor in postoperative complications and the impact of the 'second-hit' of postoperative complications on long-term survival after curative cancer surgery.
